Lake Photic Zone Temperature accross the Conterminous United States

### 2019-05-06 manuscript/manuscript.Rmd
Adding text to intro and methods

During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world (Williamson et al. 2009, Schindler 2009).  As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)(NASA GISS), Surface temperatures of lakes are also increasing globally (0.34 per decade from 1985 to 2009) [@o2015rapid].  The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.  Ultimately, temperature changes will greatly impact all aspects of lake resource management.  For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over other phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxin cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.   This need for accurate lake temperature forecasting is will become crucial for protecting human and environmental health.

Therefore, it is not surprising that modelling near-surface lake temperature has been broadly researched.  These studies typically vary in number of lakes studied, complexity of modelling approach, and time interval.  Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).  Such studies include: [@saeed2016water; @peeters2002modelin; @zhong2016recent].  There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].  By and large, air temperature and lake size are often the only selected predictor variables.  Finally, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.  Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].  Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.  Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes (Downing et al. 2006, Winslow et al. 2014).

@@

>>>>

During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].  As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)(NASA GISS), Surface temperatures of lakes are also increasing globally (0.34 per decade from 1985 to 2009) [@o2015rapid].  The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.  Ultimately, temperature changes will greatly impact all aspects of lake resource management.  For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over other phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxin cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.   This need for accurate lake temperature forecasting is will become crucial for protecting human and environmental health.

Therefore, it is not surprising that modelling near-surface lake temperature has been broadly researched.  These studies typically vary in number of lakes studied, complexity of modelling approach, and time interval.  Modelling efforts include efforts to model a single lake predicted over relatively small-time intervals (like hourly predictions).  Such studies include: [@saeed2016water; @peeters2002modelin; @zhong2016recent].  There are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].  By and large, air temperature and lake size are often the only selected predictor variables.  Finally, very few modelling efforts attempt to predict lake temperature across large spatial extents for large number of lakes.  Yet some examples include: [@minns2017factors; @o2015rapid; @wan2017comprehensive].  Despite this great variation in approach and execution, most have been rather successful modelling lake photic zone temperature, although it is important to note that the majority of these lakes are large.  Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these small lakes [@downing2006global; @winslow2014lake].

Therefore, we still lack the ability to robustly model near-surface temperature for all lakes across a large spatial extent.  Modelling lake temperature requires a large amount of data, and therefore study lakes are often selected opportunistically, which may introduce a spatial bias.  Frequently, the study lakes have a high regional resource value (e.g. the Great Lakes) and commonly have an extensive monitoring history due  to the vested interest of the public.  When model efforts do attempt to cross large spatial extents, these efforts often rely on satellite data.  While these models predict over large areas, they are restricted by the size of the lake captured by satellite (typically 3km2 for 1 km MODIS pixels).

The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).  The NLA is stratified random sample of all lakes in the conterminous United State.  The NLA was conducted in 2007 and then repeated in 2012.  Even though this is a large effort involving numerous agencies, the sampling methods are standardized and have a comprehensive quality assurance plan.  The uniqueness of this data set allows us to build a robust lake photic zone temperature model for all US lakes.

The modelling effort presented here uses the US Environmental Protection Agency’s (EPA’s) National Lake Assessment (NLA).  The NLA is stratified random sample of all lakes in the conterminous United States.  The NLA was conducted in 2007 and then repeated in 2012.  Even though this is a large effort involving numerous agencies, the sampling methods are standardized and have a comprehensive quality assurance plan.  The uniqueness of this data set allows us to build a robust lake photic zone temperature model for all US lakes.  By using lakes across the US (i.e. at a large spatial extent), we included lakes with different morphologies, in different climates with diverse geologies, and surrounding landscapes.  Additionally, small lakes were well represented with more than 50% of the lakes in the 2012 survey less than 0.5 km2.

PARAGRAPH about testing hypotheses

In summary, the main goal of this work was to develop a simple yet robust lake photic temperature model for all lakes in the conterminous US while still capturing key drivers of near-surface lake temperature. Additionally, this work (something about open science and GitHub link).

@@

##Data
We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment (NLA) (CITE).  The NLA is a stratified random sample of lakes (great than 1 ha) across the United States.  NLA sampling took place in 2007, 2012, and 2017.  This research effort used the 2007 and 2012 sample years.  The 2017 data are currently undergoing quality control before being released the public.   For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (Figure 1).  For each sampled lake, we used the mean temperature for all sampled depths of less than 2m (this being the depth typically considered the photic zone).

Figure 1: Map of 2007 and 2012 US EPA’s National Lake Assessment (NLA) Lakes

In addition to climate, there are additional factors clearly impacting lake temperature (e.g., surrounding land use, lake depth, size, and configuration, and elevation).   To test for the relative importance of lake morphometry and surrounding landscape, we used the R packages lmorpho and elevatr (CITE) as means to incorporate these types of data. Additionally, National Land Cover Database (NLCD) was our source for landcover data.  Specifically, we calculated the percent impervious surface of a 3000m buffer for each lake.  Lakes with partials buffers falling outside the US were excluded.

As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).  This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf).  PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge (Daly et al 2008).  For our study, we used the “prism” R package (Hart & Bell) to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012.

To test the relative influence of both short and long-term temperature, we derived several measures for a lake’s local air temperature.  Mean air temperatures for day of and the day before the sample date were extracted directly from the prism data.  To understand longer term influences we calculated average mean air temperatures for periods 3, 7, and 30 days prior to the sample date.

##Random Forest Modelling

Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and measure of relative variable importance.  Random forest modelling is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees).  Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.  All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).  Model performance is reported as mean square error and adjusted R.  Additionally, we used percent increase of mean-squared error to assess variable importance.  The percent increase in mean-squared error is a comparison between the true values of a variable and randomly permuted values of a specific variable.
